Mathematical Aspects of the Capital Market

### 2014-05-10 thesis/thesis.tex
Revision following Sven's advice.

% Title
%\HRule \\[0.4cm]
{ \huge \bfseries Returns Models and Their Implications
  on Cross-Correlation Matrix \\[0.4cm] }

@@

Mathematical models that describe the aforementioned observables in
terms of probability distributions, autocorrelations and cross
correlations give predictive power and help determine a fair price of
a given asset. Hence they are a central topic of the mathematics of
the financial market.

Once a model has been tentatively constructed, it must be
verified. The verification again relies on the statistics of the
observables. Over the years, a number of phenomena have been
consistently reported and become generally accepted. These are referred
to as stylized facts, which are detailed in \S
\ref{sec:StylizedFacts} and constitute a standardized test suit for
model verification.

@@

However, it is understood that real stock/index returns are much more
complicated than a straight-forward Gaussian or L\'evy distribution can
describe - instead, one needs structured models for their description.

Therefore in this thesis we study how the elements and the eigenvalues
of the cross-correlation matrix are distributed when the matrix is
built from returns described by different returns' models, specifically
GARCH(1,1) and ARIMA-log-volatility models, whose the tail behavior is
understood. Moreover, we also study how auto-correlations in the
returns influence the aforementioned distributions -- such
auto-correlations, known as second-order auto-correlations decay
exponentially but may still leave footprints in the cross-correlation
matrix.

The rest of this report is organized as follows: in
\S\ref{sec:FundamentalConcepts} we present a few very important
concepts and notations that we will often refer to in later
chapters. Then chapter \ref{chp:PriceModels} reviews some of the most
influential returns' models, calibrates them to a few intra-day return
series and compares their performance. The unconditional distribution
functions of ARIMA-log-volatility models, which is not seen to be
given a full discussion in the literature, are also calculated and
discussed. Chapter \ref{chp:Gaussian} investigates the elements' and
eigenvalues' distributions of the Wishart matrix, where the influence
of auto-correlations in the constituent returns are studied in
detail. Chapter \ref{chp:CrossCorrelationFat} looks at the elements'
and eigenvalues' distributions

\section{Fundamental Concepts \&
  Notations}\label{sec:FundamentalConcepts}
This section is a list of a few concepts that may be unfamiliar to the
reader and that we will often refer to later in the thesis:
\begin{itemize}
\item Return. Given a fixed time interval $[t - \Delta t, t]$, for example, a
  day, a week, a month, etc, and a particular asset, for example, a
  share in company ABC, the return of this asset over the time
  interval is defined as
  \begin{eqnarray*}
    r_{t, \Delta t} &=& \ln p_{t} - \ln p_{t - \Delta t} \\
    &=& \ln \left(1 + {p_{t} - p_{t - \Delta t} \over p_{t - \Delta t}}\right) \\
    &\approx& {p_{t} - p_{t - \Delta t} \over p_{t - \Delta t}}
  \end{eqnarray*}
  where $p_t$ is the price of the asset at time $t$. $\Delta t$ is
  sometimes called the time-lag of the return. Quite often, where
  confusion is not possible, we will just write $r_t$ to mean $r_{t,
    \Delta t}$, the time lag either does not matter or is clear from
  the context.

\item Autocorrelation. By autocorrelation, denoted $\rho_k$ here, we
  mean the correlation between two temporally separated observations
  of the same time series:
  \begin{eqnarray*}
    \rho_k &=& {
      \E\left[(a_t -\E(a_t))(a_{t-k} - \E(a_{t-k})\right]
      \over
      \sqrt{\text{var}(a_t)}\sqrt{\text{var}(a_{t-k})}
    }
  \end{eqnarray*}
  Here $\E(x)$ stands for the expectation value of $x$, and
  $\text{var}(x)$ stands for the variance of $x$. $k$ is called the
  time-lag and is the temporal seperation of the two observations
  measured by the number of observations in between. For example, the
  time-lag between the 1st and 3rd observation is 2.

\item Cross-Correlation Matrix. Correlations between the returns of a
  group of assets are described by the cross-correlation matrix. When
  the asset returns are described by a stable distribution law with
  L\'evy index $\alpha$, the empirical cross-correlation matrix is
  constructed as
  \begin{equation}
    \label{eq:cross-correlation}
    \begin{aligned}
      C_{ij} &= {1 \over T^{2/\alpha}} \sum_{t=1}^T [r_{i,t}-\E(r_i)]
      [r_{j,t}-\E(r_j)] \\
      &= {1 \over T^{2/\alpha}} RR'
    \end{aligned}
  \end{equation}
  where $r_{i,t}$ is the return of the i-th asset at time t and is
  placed at the entry (i, t) of matrix R; $R'$ denotes the
  transpose of R. In most practical situations, one has abundant data
  for each and every asset. Thus $T \geq N$ is assumed throughout this
  thesis.

\item Auto-regressive processes. A time series $r_t$ is called an
  auto-regressive process of order $p$ and denoted AR(p), if it can be
  written in the following form:
  \begin{eqnarray*}
    r_t &=& \sum_{i=1}^p \phi_i r_{t-i} + a_t
  \end{eqnarray*}
  where, for all $i$, $\phi_i \in (-1, 1)$, and the $a_t$'s are
  independent and identically distributed (iid.) random variables with
  zero mean. Obviously this implies the mean of $r_t$ is zero
  too. Apart from this, their distribution of $a_t$ is not restricted
  to any particular form.

  Of particular interest to this thesis is the AR(1) process:
  \begin{eqnarray*}
    r_t = \phi r_{t-1} + a_t
  \end{eqnarray*}
  Its autocorrelation function $\rho_k = \text{corr}(r_t, r_{t-k})$
  ($k = 0, 1, 2, \cdots$) can be easily shown to fall off
  exponentially:
  \begin{eqnarray*}
    \rho_k &=& {\E(r_tr_{t-k}) - \E(r_{t-k})\E(r_{t}) \over
      \sqrt{\var(r_t) \var(r_{t-k})}} \\
    &=& {\phi \E(r_{t-1}r_{t-k}) + \E(a_t r_{t-k})
      \over
      \sqrt{\var(r_t) \var(r_{t-k})}
    } \\
    &=& \phi \rho_{k-1}
  \end{eqnarray*}
  where $\E(a_t r_{t-k}) = 0$ follows from the fact that any return
  $r_{t1}$ must not depend on disturbances $a_{t2}$ that occur later
  in time. The last equation means $\rho_k$ is a geometric series. Since
  $\rho_0 = 1$, we have
  \begin{eqnarray*}
    \rho_k &=& \phi^k \\
    |\rho_k| &=& e^{k\ln|\phi|} \\
  \end{eqnarray*}
  Note that $|\phi| < 1$ and hence $\ln|\phi| < 0$.

  Although $k$ can only take integer values, it is still useful to
  define a correlation time $\tau$ such that $\phi^\tau = 1/2$. Such a
  quantity is more intuitive and constitutes a measure of
  autocorrelations that is universal and comparable among different
  time series' models. From the definition of $\tau$ we get
  \begin{equation}
    \label{eq:tau_def}
    \begin{aligned}
      \phi^\tau &= 1/2 \\
      \tau &= -{\ln 2 \over \ln\phi} \\
      \phi &= 2^{-1/\tau}
    \end{aligned}
  \end{equation}
  Figure \ref{fig:AR1-autocorrelation} shows the autocorrelation
  function of the AR(1) model. As proven above, this function decays
  exponentially.
  \begin{figure}[htb!]
    %\vspace{-15mm}
    \centering
    \includegraphics[scale=0.5, clip=true, trim=113 229 115
    139]{../pics/AR1-autocorrelation.pdf}
    \caption{\small \it Autocorrelation function of the AR(1) model
      with $\phi=1/\sqrt{2}$. Red circles: autocorrelations at $k=0, 1,
      2, \cdots$. Blue line: $e^{t\ln\phi}$.}
    \label{fig:AR1-autocorrelation}
  \end{figure}
  For more details about conventional time series' models, see
  \cite{BoxJenkins94}.
\end{itemize}

For purposes of later reference, we also list some notations that may
cause confusion to the reader:
\begin{itemize}
\item $\E(x)$ or $\mean{x}$: The expectation value of $x$.
\item $\var(x)$: the variance of $x$.
\item $\text{cov}(x,y)$: the covariance of $x$ and $y$.
\item $\text{corr}(x,y)$: the correlation between $x$ and $y$, i.e.
  \begin{equation*}
    \text{corr}(x,y) = {\text{cov}(x,y) \over \sqrt{\var(x)\var(y)}}
  \end{equation*}
\end{itemize}

\section{Stylized facts}\label{sec:StylizedFacts}
This section presents and explains a few ``stylized facts'',
i.e. phenomena that are widely observed and accepted as true.
\begin{itemize}
\item Fat tails. It has been consistently reported by various studies
  - for example \cite{Potters2003} and \cite{Mantegna2000} - that the
  probability density function (PDF) of stock/index returns are not
  Gaussian. Unlike the Gaussian PDF, which is symmetric and falls off
  very quickly as its argument moves from the center to the outskirts
  (tails), the PDF of stock/index returns are higher on the tails,
  i.e. the probability of large fluctuations is higher than is dictated
  by a Gaussian distribution - in fact, even higher than dictated by
  an exponential function. Empirical studies suggest the distribution
  function of the returns follows a power-law on the tails --- the
  exponent of the power depends on the specific stock/index.

  Figure \ref{fig:FatTail} illustrates this feature.
  \begin{figure}[htb!]
    \centering
    \includegraphics[scale=0.6, clip=true, trim=92 229 110
    140]{../pics/FatTail.pdf}
    \caption{\small \it Fat tails of {\it Nordea Bank} 15-minute
      returns during the period 2013-10-10 and 2014-01-29. The returns
      are computed using minute-by-minute average prices.}
    \label{fig:FatTail}
  \end{figure}

\item Non-zero skewness. Apart from fat tails, the PDF of stock/index
  returns are often also skewed. If the skewness is positive
  (negative), the probability of very large positive (negative)
  returns is higher than that of very large negative (positive)
  returns, even though the mean of the returns is 0 or extremely
  close to 0.

  Table \ref{tab:EmpiricalSkewness} lists the skewness of a few
  Swedish stocks traded on the Stockholm OMX market.
  \begin{table}[htb!]
    \footnotesize
    \centering
    \begin{tabular}{|c|c|c|c|c|c|c|}
      \hline
      Nordea Bank & Volvo B & Boliden & ABB Ltd & H\&M & Scania
      B & Ericsson B \\
      \hline
      0.1362 & 0.0471 & 0.0567 & -0.0364 & -0.0168 &
      -1.1554 & 0.2086 \\
      \hline
    \end{tabular}
    \caption{\small \it Skewness of Stock Returns. All the returns
      have time-lag of 15 minutes and are computed using paid prices
      between 2013-10-10 and 2014-01-29. }
    \label{tab:EmpiricalSkewness}
  \end{table}

\item Higher-order autocorrelation. To the lowest order, return series
  are not auto-correlated --- if they are, the auto-correlations would
  present an obvious opportunity of making easy profit and be
  exploited and vanish as soon as they appear. However, the squared
  returns do have significant auto-correlations, as figure
  \ref{fig:nordea_15min_acf} illustrates. These auto-correlations
  suggest the variances of the returns at subsequent time steps are
  correlated. These are referred to as higher-order auto-correlations
  and make a subjet of returns' models.

\end{itemize}

@@

\input{summary}
\input{appendix}

>>>>

% Title
%\HRule \\[0.4cm]
{ \huge \bfseries Returns Models and Their Covariance Matrix
  \\[0.4cm] }

@@

Mathematical models that describe the aforementioned observables in
terms of probability distributions, autocorrelations and cross
correlations give predictive power and help determine a fair price of
a given asset. Hence they are a central topic of the mathematics of
the financial market. In the domain of discrete-time returns' models,
GARCH and stochastic volatility models (SV) are the most popular, so in
this thesis project we study a total of 7 intraday return series 
using GARCH and SV models and compare their performance in volatility
forecast.

% Once a model has been tentatively constructed, it must be
% verified. The verification again relies on the statistics of the
% observables. Over the years, a number of phenomena have been
% consistently reported and become generally accepted. These are referred
% to as stylized facts, which are detailed in \S\ref{sec:StylizedFacts}
% and constitute a standardized test suit for model verification.

@@

However, it is understood that real stock/index returns are much more
complicated than a straight-forward Gaussian or L\'evy distribution can
describe --- instead, one needs structured models for their description.

Therefore in this thesis we study how the elements and the eigenvalues
of the cross-correlation matrix are distributed when the matrix is
built from returns described by different returns' models, specifically
GARCH(1,1) and stochastic volatility models, whose the tail behavior is
understood. Moreover, we also study how auto-correlations in the
returns influence the aforementioned distributions --- such
auto-correlations, known as second-order auto-correlations decay
exponentially but may still leave footprints in the cross-correlation
matrix.

The rest of this report is organized as follows:

% \S\ref{sec:FundamentalConcepts} presents a few very important
% concepts and notations that will often be referred to in later
% chapters; \S\ref{sec:StylizedFacts} gives a review of the ``stylized
% facts'', i.e. a number of consistently reported and generally accepted
% phenomena that must be accounted for by a returns' model.

Chapter \ref{chp:PriceModels} reviews some of the most
influential returns' models, fits them to a few intraday return
series and compares their performance. A calculation of the
unconditional distribution functions of stochastic volatility models,
especially in the case where the residual of the log-volatility
and the innovation of the return are correlated normal variates, is
also presented. Chapter \ref{chp:Gaussian} investigates the elements' and
eigenvalues' distributions of the Wishart matrix, and studies the influence
of auto-correlated returns. Chapter \ref{chp:CrossCorrelationFat}
looks at the elements' and eigenvalues' distributions of a covariance
matrix that is built from identically specified GARCH(1,1) series.

@@

\input{summary}
\appendix
\input{appendix}
